---
title: "ecoPointXAI: Explainable 3D Forest Analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecoPointXAI Intro Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  warning = FALSE,
  message = TRUE
)
```

# 🧩 ecoPointXAI Introduction

This vignette provides a compact, end-to-end example of how to use **ecoPointXAI**
for explainable 3D forest analytics — from voxel data preprocessing to model explainability visualization.

---

## 1️⃣ Load required libraries

```{r}
library(ecoPointXAI)
library(torch)
library(rgl)
```

---

## 2️⃣ Prepare example voxel data

For demonstration, the function `example_voxel_block()` generates a small 3D voxel sample (X, Y, Z, value).

```{r}
# Try using example_voxel_block() if available, else create a demo dataset
if (exists("example_voxel_block", mode = "function")) {
  vox <- example_voxel_block()
} else {
  set.seed(123)
  vox <- data.frame(
    X = runif(200, 0, 10),
    Y = runif(200, 0, 10),
    Z = runif(200, 0, 5),
    value = runif(200)
  )
}

summary(vox)
hist(vox$Z, main = "Voxel Heights Distribution", col = "forestgreen", border = NA)
```

---

## 3️⃣ Train a minimal PointNet model

We define and train a lightweight model on the example dataset for just a few epochs.

```{r, eval=FALSE}
model <- pointnet_model(input_dim = 4, output_dim = 1)

trained <- train_pointnet(
  model,
  train_loader = vox$train,
  val_loader   = vox$val,
  epochs       = 5,
  patience     = 2
)
```

---

## 4️⃣ Generate explainability maps

Once trained, the model can be explained using either gradient-based or SHAP methods.
Here, we use **gradient-based** voxel importance.

```{r, eval=FALSE}
grad_exp <- explain_pointnet(
  model       = trained$model,
  voxel_data  = vox$features,
  method      = "gradients"
)

head(grad_exp$feature_importance)
```

---

## 5️⃣ Visualize explainability in 3D

We can now visualize voxel-level importance fields interactively in RGL or export them as an HTML scene.

```{r, eval=FALSE}
visualize_explainability(
  vox_df         = vox$coords,
  explain_result = grad_exp,
  export_path    = tempfile(fileext = ".html")
)
```

This will open an interactive 3D viewer showing the **importance distribution** of each voxel.

---

## ✅ Summary

In this vignette, you have:

1. Loaded voxel-based LiDAR data  
2. Trained a minimal PointNet model  
3. Computed gradient-based explainability  
4. Visualized model attention in 3D space  

To explore advanced visualization (e.g., **virtual forest rendering**, SHAP comparison, or ecological interpretation),
see the companion vignette `vignettes/virtual_forest.Rmd`.

---

## 🔗 Session Info

```{r}
sessionInfo()
```
